---
title: "packedMatrix"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Based on discussion on R-devel beginning
[here](https://stat.ethz.ch/pipermail/r-devel/2021-November/081261.html)
and continued privately with Martin Mächler.

## Class definition

```{r showClass}
library("Matrix")
showClass("packedMatrix")
```

## Utilities for tests

`checkEquiv` takes as an argument a list of calls `FUN(x, ...)` 
and returns a logical vector of the same length, indicating whether 
the values of `FUN(x, ...)` are consistent with the values of
`FUN(as(x, "matrix"), ...)`. (Here, _consistent_ means identical 
modulo a coercion from `"Matrix"` to `"matrix"`.)

```{r utils1}
checkEquiv <- function(exprs) {
  f <- function(x1, envir) {
    stopifnot(is.call(x1), length(x1) > 1L)
    x2 <- x1
    x2[[2L]] <- call("as", x2[[2L]], "matrix") 
    m1 <- eval(x1, envir)
    m2 <- eval(x2, envir)
    if (is(m1, "Matrix")) {
      m1 <- as(m1, "matrix")
    }
    identical(m1, m2)
  }
  res <- vapply(exprs, f, NA, parent.frame())
  names(res) <- vapply(exprs, deparse, "")
  res
}
```

`checkTimes` takes as an argument a list of calls `FUN(x, ...)` 
and returns a benchmark comparing their evaluation times to
those of the calls `FUN(as(x, "matrix"), ...)`.

```{r utils2}
checkTimes <- function(exprs, times = 100L) {
  stopifnot(requireNamespace("microbenchmark"))
  f <- function(x) {
    stopifnot(is.call(x), length(x) > 1L)
    x[[2L]] <- call("as", x[[2L]], "matrix")
    x
  }
  l <- vector("list", 2L * length(exprs))
  l[c(TRUE, FALSE)] <- exprs
  l[c(FALSE, TRUE)] <- lapply(exprs, f)
  l <- c(list(quote(microbenchmark::microbenchmark)), l, list(times = times))
  eval(as.call(l), parent.frame())
}
```

## `[` method tests

For `"packedMatrix"`, the subset operator ```[``` avoids coercion 
to `"matrix"` in many special cases. Here is a list of operations
to be tested.

```{r subset1}
exprs <- alist(S[integer(0L)],
               S[c(0:6, NA)],
               S[-(0:6)],
               S[logical(0L)],
               S[TRUE],
               S[FALSE],
               S[NA],
               S[c(TRUE, FALSE, NA)],
               S[integer(0L), ],
               S[1L, ],
               S[1L, , drop = FALSE],
               S[c(0:1, NA), ],
               S[c(0:2, NA), ],
               S[-seq_len(nrow(S)), ],
               S[-seq_len(nrow(S))[-1L], ],
               S[-seq_len(nrow(S))[-(1:2)], ],
               S[logical(0L), ],
               S[TRUE, ],
               S[FALSE, ],
               S[NA, ],
               S[c(TRUE, FALSE, NA), ],
               S[character(0L), ],
               S[rownames(S)[1L], ],
               S[rownames(S)[1L], , drop = FALSE],
               S[rownames(S)[1:2], ],
               S[matrix(0L, 0L, 2L)],
               S[cbind(c(0:6, NA), c(NA, 6:0))],
               S[cbind(c(rownames(S), NA), c(NA, colnames(S)))],
               S[matrix(c(TRUE, FALSE), nrow(S), ncol(S))],
               S[NULL],
               S[NULL, ],
               S[, NULL])
```

We start by checking that the `"packedMatrix"` subset is consistent
with the subset via `"matrix"`.

```{r subset2}
n <- 10L
S <- new("dspMatrix", uplo = "U", x = as.double(seq_len(choose(n + 1L, 2L))),
         Dim = c(n, n), Dimnames = list(A = sprintf("i%d", seq_len(n)), B = NULL))
all(checkEquiv(exprs))
```

Now we test its performance against the subset via `"matrix"`.

```{r subset3, echo=-1}
options(width = 500L)
n <- 1000L
S <- new("dspMatrix", uplo = "U", x = as.double(seq_len(choose(n + 1L, 2L))),
         Dim = c(n, n), Dimnames = list(A = sprintf("i%d", seq_len(n)), B = NULL))
checkTimes(exprs)
```

The `"packedMatrix"` subset is slower in some cases because:

* It is implemented in R rather than C (for now).
* It takes extra steps to avoid a coercion to `"matrix"`.

However, for large enough `n`, there is no comparison...

```{r subset4, error=TRUE}
n <- 30000L
i <- replace(logical(1000L), 1L, TRUE)
S <- new("dspMatrix", uplo = "U", x = as.double(seq_len(choose(n + 1L, 2L))), Dim = c(n, n))
{S[i]; cat("done!\n")}
{as(S, "matrix")[i]; cat("done!\n")}
```

## `t` method tests

```{r transpose1}
selectMethod("t", signature(x = "packedMatrix"))
n <- 1000L
L <- new("dtpMatrix", uplo = "L", x = as.double(seq_len(choose(n + 1L, 2L))), 
         Dim = c(n, n), Dimnames = list(A = NULL, B = sprintf("j%d", seq_len(n))))
identical(as(t(L), "matrix"), t(as(L, "matrix")))
microbenchmark::microbenchmark(t(L), as(t(as(L, "dtrMatrix")), "dtpMatrix"))
```
```{r transpose2, error=TRUE}
n <- 30000L
L <- new("dtpMatrix", uplo = "L", x = as.double(seq_len(choose(n + 1L, 2L))), Dim = c(n, n))
{t(L); cat("done!\n")}
{as(t(as(L, "dtrMatrix")), "dtpMatrix"); cat("done!\n")}
```

## `diag` method tests

```{r diag1}
selectMethod("diag", signature(x = "packedMatrix"))
n <- 4L
U <- new("lspMatrix", uplo = "U", x = rep_len(c(TRUE, FALSE, NA), choose(n + 1L, 2L)),
         Dim = c(n, n), Dimnames = list(letters[seq_len(n)], NULL))
U
diag(U)
diag(U, names = FALSE)
```

## `diag<-` method tests

```{r diag2, error=TRUE}
selectMethod("diag<-", signature(x = "packedMatrix"))
U
diag(U) <- logical(n)
U
diag(U) <- TRUE
U
diag(U) <- 1 # 'x@x' coerced to 'typeof(value)'
U
diag(U) <- TRUE # 'value' coerced to 'typeof(x@x)'
U
diag(U) <- double(2L) # right type, wrong length
diag(U) <- vector("list", n) # right length, wrong type
```
